
```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.path = "man/figures/",
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)
```

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/CYGUBICKO/varpred.svg?branch=main)](https://travis-ci.com/CYGUBICKO/varpred)
[![CRAN status](https://www.r-pkg.org/badges/version/varpred)](https://CRAN.R-project.org/package=varpred)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->


This package implements two approaches for constructing outcome plots (prediction and effect plots). These include:

- **mean-based** approach
- **observed-value** approach

It can also be used to generate bias-corrected prediction and effect estimates for generalized linear models involving non-linear link functions, including models with random effects. This package complements the existing ones by providing:

- a straightforward way to generate effects plots
- a robust way to correct for non-linear averaging bias in generalized (mixed) models


## Installation

You can install the development version of varpred from [GitHub](https://github.com/cygubicko/varpred) with:

```{r install, eval=FALSE}
# install.packages("remotes")
remotes::install_github("CYGUBICKO/varpred")
```

## Example

We use `mtcars` data to show outcome plots:

- `isolate=TRUE` to generate effect plot
- `isolate=FALSE` to generate prediction plot

```{r simple_example}
library(varpred)
library(ggplot2)

## Set theme for plots
varpredtheme()

## Fit the model
mod <- lm(mpg ~ wt + hp, mtcars)

## Effect
ef <- varpred(mod, "wt", isolate=TRUE, modelname="effect")
plot(ef)

## Prediction
pred <- varpred(mod, "wt", isolate=FALSE, modelname="prediction")
plot(pred)

## Compare effect and prediction
all_v <- combinevarpred(list(ef, pred))
p1 <- plot(all_v)

## Add observed data
print(p1
	+ geom_point(data=mtcars, aes(x=wt, y=mpg), col="grey")
	+ labs(colour="Method", linetype="Method")
)
```

