---
title: "Outcome plots: uncertainty estimation and bias correction for predictions and effects in simple and generalized linear (mixed) models"
author: Bicko Cygu, Ben Bolker, Jonathan Dushoff
output:
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Outcome plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

\def\code#1{\texttt{#1}}
\let\proglang=\textsf

# Motivation

The development of this package is motivated by the water, sanitation, and hygiene (WaSH) data in which we were interested in investigating the contribution of demographic and socio-economic factors to improved WaSH indicators among the slum dwellers in Nairobi, Kenya. We noticed that the predictions we generated using the existing packages consistently over- or under- estimated the observed proportions; and did not align well with the observed data points. In other words, what we call _bias_. There are several (challenges) reasons for this, including:

- the choice of the _reference point_
- uncertainty estimation -- the choice of _anchor_ for computing confidence intervals
- biases induced by non-linear averaging due to non-linear transformation in generalized linear models

# Introduction

An outcome plot (often called effect or prediction plot), plots _focal_ predictor on the x-axis and _central estimate_ (predicted values) on the y-axis. The resulting plot provide a useful way to summarize the results from a regression model; and depend on how we deal with the above challenges.

## Mean-based approach

In a model with non-focal predictors such as multivariate models, reference points (values chosen for non-focal predictors) can be chosen as the average of the non-focal _linear predictor variables_ (columns of the model matrix corresponding to non-focal predictors) -- we call this approach _mean-based_ reference point and is currently not implemented in commonly used \proglang{R} software \textbf{packages}.

```{r setup}
library(varpred)
```
